---
title: Homework 07
author: Sriram Krishnaswamy
---

# Problem 1
Since previous studies have reported that elite athletes are often
deficient in their nutritional intake, a group of researchers decided
to evaluate Canadian high performance athletes. A total of $n = 324$
athletes from eight Canadian sports centers participated in the study.
One reported finding was that the average caloric intake among the $n
= 201$ women was 2403.7 kcal/day. The recommended amount is 2811.5
kcal/day. Is there evidence that female Canadian high performance
athletes are deficient in caloric intake?

1. State the appropriate $H_0$ and $H_A$ to test this.
2. Assuming a population standard deviation of 880 kcal/day, carry out
   the test at the $\alpha$ = 0.05 level. Make sure to interpret the
   results in plain language.

   Now suppose a nutritionist is brought in to implement a new health
   program for the female athletes. This program should increase mean
   caloric intake but not change the standard deviation. Given the
   standard deviation, and how calorically deficient these athletes
   are, a change in the mean of 50 kcal/day to 2453.7 is of little
   practical importance. However, with a large enough sample, this
   change can be statistically significant. To see this, calculate the
   P-value for the test of ${H}_{0} : \mu = 2403.7$ versus $H_A : \mu
   > 2403.7$. in each of the following situations:

3. In a sample of 500 athletes, the average caloric intake is $\bar{x}$ = 2453.7.
4. In a sample of 2500 athletes, the average caloric intake is $\bar{x}$ = 2453.7.

## Solution

### Part 1

The null hypothesis and alternative hypothesis are given by

\[
{H}_{0} : \mu_0 = 2811.5 \text{ kcal/day}
\]

\[
{H}_{A} : \mu < \mu_0 \hspace{0.25cm} \text{or} \hspace{0.25cm} \mu < 2811.5 \text{ kcal/day}
\]

### Part 2

Given $\sigma = 880$ kcal/day and $\alpha = 0.05$. For the sample with
women, $\bar{x} = 2403.7$ and $n = 201$. First, we calculate the
*z*-score of the population.

\[
z = \frac{\bar{x} - \mu}{\frac{\sigma}{\sqrt{n}}} = \frac{2403.7 -
2811.5}{\frac{880}{\sqrt{201}}} = \frac{-407.8}{62.07} = -6.57
\]

For the alternate hypothesis, ${H}_{A} : \mu < 2811.5$, we need to
calculate the *p*-value.

\[
p = P[z \leq -6.57]
\]

Since there are no tables for such an extreme value, we calculate it
using R.

```{r}
pnorm(-6.57)
```

As it can be seen, the value is $\approx 0$. Therefore, we can
conclude $p \approx 0$. At level $\alpha = 0.05$, we check $0.0 <
0.05$ is true and *reject* the null hypothesis that $\mu = 2811.5$ in
favor of the alternative hypothesis ${H}_{A} : \mu < 2811.5$ because
$\bar{x} = 2403.7 < 2811.5$. In plain words, this test provides
evidence that there is a deficiency in caloric intake among female
Canadian high performance athletes and the sample is not just an
outlier.

### Part 3
For this sample, $n = 500$ and $\bar{x} = 2453.7$. The standard
deviation doesn't change and, therefore, is $\sigma = 880$ kcal/day.
Our hypothesis test becomes

\[
{H}_{0} : \mu = 2403.7
\]

\[
{H}_{A} : \mu > 2403.7
\]

Now, we calculate the *z*-score

\[
z = \frac{\bar{x} - \mu}{\frac{\sigma}{\sqrt{n}}} = \frac{2453.7 -
2403.7}{\frac{880}{\sqrt{500}}} = \frac{50}{39.35} = 1.27
\]

Then, we calculate the corresponding *p*-value

\[
p = P[z \geq 1.27] = 1 - P[z < 1.27]
\]

We can perform this calculation using R.

```{r}
1 - pnorm(1.27)
```

Therefore, $p = 0.1$. If we take it at the same level $\alpha = 0.05$,
then we have $0.1 > 0.05$ and we keep the null hypothesis that $\mu =
2403.7$ kcal/day. In other words, a sample of 500 athletes indicating
an average 50 kcal/day increase is not statistically significant to
validate the new strategy by the nutritionist.

### Part 4

For this sample, $n = 2500$ and $\bar{x} = 2453.7$. The standard
deviation doesn't change and, therefore, is $\sigma = 880$ kcal/day.
Our hypothesis test becomes

\[
{H}_{0} : \mu = 2403.7
\]

\[
{H}_{A} : \mu > 2403.7
\]

Now, we calculate the *z*-score

\[
z = \frac{\bar{x} - \mu}{\frac{\sigma}{\sqrt{n}}} = \frac{2453.7 -
2403.7}{\frac{880}{\sqrt{2500}}} = \frac{50}{17.6} = 2.84
\]

Then, we calculate the corresponding *p*-value

\[
p = P[z \geq 2.84] = 1 - P[z < 2.84]
\]

We can perform this calculation using R.

```{r}
1 - pnorm(2.84)
```

Therefore, $p = 0.0022$. If we take it at the same level $\alpha =
0.05$, then we have $0.0022 < 0.05$ and we *reject* the null
hypothesis that $\mu = 2403.7$ kcal/day. In other words, a sample of
2500 athletes indicating an average 50 kcal/day increase is
statistically significant to validate the new strategy by the
nutritionist.

# Problem 2
Thirty samples were taken from a stream, and the pollution level in
parts per million (ppm) was recorded for each sample. The average
pollution level for the data was $\bar{x}$ = 10.1ppm. Suppose that the
population standard deviation is 2.7ppm. The investigator who
collected the samples is interested in determining whether or not
there is evidence that the population mean pollution level $\mu$ is
greater than 9.5ppm. The hypotheses are:

\[
{H}_{0}: \mu = 9.5 \hspace{0.25cm} \text{vs.} \hspace{0.25cm} {H}_{A} : \mu > 9.5
\]

1. Assuming $\alpha$ = 0.05, for what values $z$ of the Z-test
   statistic will you reject the null hypothesis?
2. Using $\sigma$ = 2.7 and $\mu$ = 9.5 (i.e., assuming ${H}_{0}$ is
   true), for what values of $\bar{X}$ will you reject ${H}_{0}$ at
   $\alpha = 0.05$?
3. Using $\sigma$ = 2.7 and $\mu$ = 10, what is the probability that
   the statistic $\bar{X}$ will fall in the region defined in part 2?
   This probability is the power of the test when $\mu$ = 10.
4. Find the power of the test when $\mu$ = 11 and compare it with your
   answer in part 3.

## Solution
We are given that $\bar{x} = 10.1$ ppm and $\hat{S} = 2.7$ ppm with $n = 30$.

\[
{H}_{0} : \mu = 9.5
\]

\[
{H}_{A} : \mu > 9.5
\]

### Part 1
We are going to reject the hypothesis if our region is in the upper
tail of 0.05. In other words, if our *p*-value is less than 0.05.

\[
P[Z \geq z] < 0.05
\]

We can calculate from the table that ${z}_{0.05} = 1.645$. Our
rejection region would be any z > 1.645. If we were doing a two-sided
test, we have to calculate $\frac{\alpha}{2}$ in both directions.

### Part 2
From our previous part, we know that *z* > 1.645 for us to reject

\[
\frac{\bar{X} - \mu}{\frac{\sigma}{\sqrt{n}}} > 1.645
\]

\[
\implies \frac{\bar{X} - 9.5}{\frac{2.7}{\sqrt{30}}} > 1.645
\]

\[
\implies \bar{X} > 1.645 \left( \frac{2.7}{\sqrt{30}} \right) + 9.5 \implies \bar{X} > 10.31
\]

Therefore, we reject our null hypothesis if our sample mean,
$\bar{X}$, is greater than 10.31 ppm.

### Part 3
We want to find the power of the test if the true mean is 10.

\[
P[\bar{X} > 10.31 | \mu = 10] = P \left[ z > \frac{10.31 - 10}{\frac{2.7}{\sqrt{30}}} \right]
\]

\[
\implies P[z > 0.63] = 1 - P[z \leq 0.63]
\]

```{r}
1 - pnorm(0.63)
```

Therefore, our *p*-value is 0.2643. We are going to correctly reject
our hypothesis around 26% of the time. To prove this, we can increase
our sample size.

### Part 4
We want to find the power of the test if the true mean is 11.

\[
P[\bar{X} > 10.31 | \mu = 11] = P \left[ z > \frac{10.31 - 11}{\frac{2.7}{\sqrt{30}}} \right]
\]

\[
\implies P[z > - 1.4] = 1 - P[z \leq - 1.4]
\]

```{r}
1 - pnorm(-1.4)
```

Therefore, our *p*-value is 0.9192. We are going to correctly reject
our hypothesis around 92% of the time. To prove this, we can increase
our sample size.

### Extra credit
This involves 2 regions for rejection.

# Problem 3
The following questions regard *t* distributions.

1. Using the *t*-distribution table, find the 95th percentile of
   $t(11)$ denoted by ${t}_{0.05}(11)$ and the 5th percentile of
   $t(11)$.
2. Find the interval containing the middle 90% of values of t
   distribution with 11 degrees of freedom and use your answer to
   provide the 90% confidence interval formula for the population mean
   $\mu$ based on a sample of size 12 with sample standard deviation
   /S/.
3. How does the 90% confidence interval formula in part 2 change if
   the sample size increases to 24?
4. How does the confidence interval formula in part 2 change if the
   confidence level increases to 95% with the same sample size of 12?
5. Suppose that the one-sample t statistic for testing $H_0 : \mu =
   10$ versus $H_A : \mu > 10$ from a sample of $n = 18$ observations
   has the value $t = 2.05$.
   - What are the degrees of freedom for this statistic?
   - Give the two critical values from the *t*-distribution table that
     bracket t and express the two values using the notation
     ${t}_{\alpha}(k)$ with appropriate $\alpha$ and $k$.
   - Between what two values does the P-value of the test fall?
   - Is the value $t = 2.05$ significant at the 5% level? Is it significant at the 1% level?
   - Using R, find the exact P-value of $t = 2.05$.

## Solution

### Part 1
From the tables,

\[
{t}_{0.05}(11) = 1.796
\]

\[
{t}_{0.95}(11) =
\]

### Part 2
The interval is given by $\bar{X} \pm t(11) \frac{S}{\sqrt{n}}$.
Substituting, the values, the interval turns out to be

\[
\bar{X} \pm {t}_{\frac{0.1}{2}}(11) \frac{S}{\sqrt{12}}
\]

\[
\bar{X} \pm 0.5185 S
\]

### Part 3
Now, our degrees of freedom has changed. $n - 1 = 23$.
${t}_{\frac{0.1}{2}}(23) = 1.714$. Our new interval is $\bar{X} \pm
1.714 \frac{S}{\sqrt{24}}$

### Part 4
Now, $n = 12$ and $\alpha = 0.05$. Our new interval is now $\bar{X}
\pm {t}_{\frac{0.05}{2}}(11) \frac{S}{\sqrt{n}}$. From the tables,
${t}_{0.025}(11) = 2.201$. Therefore, the interval becomes $\bar{X}
\pm 2.201 \frac{S}{\sqrt{12}}$, i.e., widens.

### Part 5
Here, $n = 18$ and $t = 2.05$. We also know that we are testing $H_0 :
\mu = 10$ versus $H_A : \mu > 10$. From the tables,

- Degrees of freedom = $n - 1$ = 17
- The critical values are 1.74 and 2.11. $1.74 \leq 2.05 \leq 2.11$.
  It is the same as ${t}_{0.05}(17) \leq {t}_{\alpha}(17) \leq
  {t}_{0.025}(17)$.
- The p values fall in the range (0.05, 0.025). In terms of intervals,
  $0.05 \geq p \geq 0.025$
- It is significant at 5% level as $0.05 \geq p \geq 0.025$ but not
  significant at 1% level since $0.01 < 0.025 < p < 0.05$.
- The exact value is 0.028

```{r}
1 - pt(2.05, 17)
```

# Problem 4
The National Endowment for the Humanities sponsors summer institutes
to improve the skills of high school teachers of foreign languages.
One such institute hosted 20 French teachers for 4 weeks. At the
beginning of the period, the teachers were given the Modern Language
Association’s listening test of understanding of spoken French. After
4 weeks of immersion in French, the listening test was given again.
The maximum possible score on the test is 36. To analyze these data,
we first subtract the pretest score from the posttest score to obtain
the improvement for each teacher. These 20 differences form a single
sample. Load the data set (`frenchteachers.RData`) to do the analyses
below.

1. Define the appropriate parameter $\mu$ in words precisely for
   measuring the change of the teachers’ listening skills from the
   pretest to the posttest.
2. Give an estimate of $\mu$, and provide the standard error of your
   estimate. Find the 95% confidence interval for $\mu$. What
   assumptions are needed for this confidence interval to be valid?
   Check any assumptions that you can.
3. State appropriate null and alternative hypotheses for examining the
   question of whether or not the course is associated with an
   improvement in French skills. Perform a *t*-test to test this
   hypothesis – make sure you calculate the appropriate test
   statistic, calculate a p-value, and summarize your conclusion
   (assuming an level of $\alpha = 0.05$) in the context of the
   problem.
4. Using `t.test` in R, verify your results in parts 2 and 3.
5. What would be a causal interpretation of the conclusion in part 4?
   Are you comfortable with this interpretation? Explain in a sentence
   or two.
6. Use the sign test to assess whether the summer institute is
   associated with an improvement in French listening skills. Make
   sure to state hypotheses, calculate the appropriate test statistic,
   calculate a p-value, and summarize your conclusion in the context
   of the problem.

## Solution
One set of people with 2 test scores and we are going to do a pair
test.

Let's load the data set first

```{r}
load("./frenchteachers.RData")
```

### Part 1
The appropriate parameter for measuring change is the difference in
scores between the tests given at the start of the program and the end
of it. In other words,

\[ \mu = {\mu}_{post} - {\mu}_{pre} \]

### Part 2
We assume normality of the data. Therefore, we have to first check if
this is true.

```{r, fig.height=3.5, fig.width=6.5, fig.show='asis', fig.align='center'}
par(mfrow = c(1, 2), cex = 0.5)
hist(frenchteachers$gain, main = "Histogram of gain")
qqnorm(frenchteachers$gain)
qqline(frenchteachers$gain)
```

Since this seems to follow a normal distribution, we can use the
normal distribution if the sample size is high. We can also use a
*t*-distribution since this satisfies the pre-requisites for it. To do
that, we have to first calculate the mean and standard deviation of
the sample given.

```{r}
 mean(frenchteachers$gain)
```

```{r}
sd(frenchteachers$gain)
```

We know that the degrees of freedom is $n - 1 = 19$ for this case.
From the tables, we know ${t}_{0.05}(19) = 1.729$.

\[ \text{C.I} = \bar{x} \pm {t}_{0.05}(19) \frac{S}{\sqrt{n}} = 0.95
\pm 1.729 \times \frac{1.395}{\sqrt{20}} = (0.411, 1.489) \]

Therfore, the 95% confidence interval is (0.411, 1.489).

### Part 3
The null and alternative hypotheses for this problem is given below

\[ {H}_{0} : \mu = 0 \]

\[ {H}_{A} : \mu > 0 \]

We are given that $\alpha = 0.05$. Now, let's construct the *t*-test
statistic

\[
t = \frac{\bar{x} - \mu}{\frac{S}{\sqrt{n}}} = \frac{0.95 - 0}{\frac{1.395}{\sqrt{20}}} = 3.045
\]

From the tables, for degrees of freedom 19, it's between 0.005 and
0.0025. We can get the exact value from *R*.

```{r}
1 - pt(3.045, 19)
```

The *p*-value is 0.0033.

### Part 4

```{r}
t.test(frenchteachers$gain, alternative = "greater", mu = 0)
```

### Part 5

The *p*-value from part 4 is 0.0033 < 0.05 and this means that we
*reject* the null hypothesis that $\mu = 0$ and, since the sample mean
is 0.95, the mean is 0.95. The causal interpretation from this data is
that the 4 weeks of immersion helped. The improvement is 0.95 or 2.79%
improvement. I'm not comfortable with the interpretation since no
information about how the French teachers were selected is given. If
the experiment was double-blind and randomized, then there is more
merit to the experiment. Otherwise, this is not a good experiment
setup.

### Part 6
  
\[
H_0 : p = 0.5 \text{ vs } {H}_{A} : p > 0.5
\]

To perform the sign test, we need the no. of observations that are
valid. The following piece of code finds the number of observations
that follow our requirement.

```{r}
length(which(frenchteachers$gain > 0))
```

```{r}
length(which(frenchteachers$gain == 0))
```

```{r}
length(which(frenchteachers$gain < 0))
```

In 20 trials, 13 had scored higher after the immersion, 3 had scored
lower, and 4 had scored the same $\implies {x}_{obs} = 13$. Since,
there is a tie, we delete those observations and the sample size is
now 16.

Now, we calculate the *t*-statistic

\[
t = \frac{\bar{x} - \mu}{\frac{S}{\sqrt{n}}} = \frac{0.95 - 0}{\frac{1.3945}{\sqrt{16}}} = 2.72
\]

From the tables, considering degrees of freedom of 15, we get $0.01 < p < 0.005$
and we can get the exact value from `R`.

```{r}
1 - pt(2.72, 15)
```

Because $p = 0.008 < 0.05$, we reject the null and conclude
the immersion course did help the French teachers perform better in
the test.
