---
title: Homework 08
author: Sriram Krishnaswamy
---

# Problem 1

Researchers used 7 red and 7 black cards to randomly assign 14
volunteer males with high blood pressure to one of two diets for four
weeks: a fish oil diet and a standard oil diet. The reductions in
diastolic blood pressure are available in `fishoil.Rdata`. Do the
following steps to compare the treatments.

1. Compute the averages and the sample standard deviations for each
   group separately
2. Compute the pooled estimate of standard deviation using the formula
   in Section 2.3.2
3. Compute $SE(\bar{Y}_2 - \bar{Y}_1)$ using the formula in Section
   2.3.2
4. What are the degrees of freedom associated with the pooled estimate
   of the standard deviation? What is the 97.5th percentile of the
   *t*-distribution with this many degrees of freedom?
5. Construct a 95% confidence interval for $\mu_2 - \mu_1$ using the
   formula in Section 2.3.3
6. Compute the *t*-statistic for testing equality as shown in Section
   2.3.5
7. Find the one-sided *p*-value (as evidence that the fish oil diet
   resulted in greater reduction of blood pressure) by comparing the
   *t*-statistic in part 6 to the percentiles of the appropriate
   *t*-distribution (by reading the appropriate percentile from a
   computer program or calculator).
   
## Solution

We can first check the given data.

```{r prob02data}
load('./fishoil.RData')
fishoil
```

### Means and standard deviations

For the volunteers that had a regular oil diet, the mean and standard
deviation are -1.14 and 3.18 respectively.

```{r mean01regular}
mean(fishoil[fishoil$Diet == 'RegularOil',1])
sd(fishoil[fishoil$Diet == 'RegularOil',1])
```
For the volunteers that had a fish oil diet, the mean and standard
deviation are 6.57 and 5.86 respectively.

```{r mean01fish}
mean(fishoil[fishoil$Diet == 'FishOil',1])
sd(fishoil[fishoil$Diet == 'FishOil',1])
```

### Pooled estimate

The pooled estimate of the standard deviation is given by the following formula ---

\[
{s}_{p} = \sqrt{\frac{(n_1 - 1) {s}^{2}_{1} + (n_2 - 1) {s}^{2}_{2}}{(n_1 + n_2 - 2)}}
\]

Therfore, for our current problem, $n_1 = 7$, $s_1 = 5.8554$, $n_2 =
7$ and $s_2 = 3.184785$.

\[
\implies {s}_{p} = \sqrt{\frac{(7 - 1) {5.8554}^{2} + (7 - 1) {3.184785}^{2}}{(7 + 7 - 2)}}
= \sqrt{\frac{6 \times 34.28571 + 6 \times 10.14286}{12}}
\]

\[
\implies {s}_{p} = \sqrt{\frac{266.57142}{12}} = \sqrt{22.214285} = 4.7132
\]

### Standard error

From the statistical sleuth, we have the following formula ---

\[
SE(\bar{Y}_2 - \bar{Y}_1) = {s}_{p} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}
\]

\[
\implies SE(\bar{Y}_2 - \bar{Y}_1) = 4.7132 \sqrt{\frac{2}{7}} = 4.7132 \times 0.53452
= 2.5193
\]

### Degrees of freedom

\[
\text{d.f.} = n_1 + n_2 - 2 = 12
\]

For getting 97.5th, $\alpha = 1 - 0.975 = 0.025$. From the tables,

\[
{t}_{0.025}(12) = 2.179
\]

### Confidence interval

From section 2.3.3 in the statistical sleuth, we get the following
formula (for $\alpha = 0.05$)

\[
\text{C.I.} = (\bar{Y}_2 - \bar{Y}_1) \pm {t}_{0.05}(12) SE(\bar{Y}_2 - \bar{Y}_1)
\]

\[
\implies \text{C.I} = 6.571429 - (-1.142857) \pm 1.782 \times 2.5193
= 7.714286 \pm 4.4893926 = (3.22489, 12.2037)
\]

Therefore, the 95% confidence interval is (3.22489, 12.2037).

### *t*-statistic

The hypothesis is that the fish oil diet does not change one's
diastolic blood level. Hence, $D = 0$. Therefore, the *t*-statistic is
given by ---

\[
t = \frac{(\bar{Y}_2 - \bar{Y}_1) - D}{SE(\bar{Y}_2 - \bar{Y}_1)} = \frac{7.714286}{2.5193}
= 3.0621
\]

### *p*-value

For the fish oil diet to be effective, $\mu_2 - \mu_1 > D$ and since
$D = 0$, $\mu_2 - \mu_1 > 0$. From the tables, the *p*-value is approximately 0.005. We can find
the exact value from `R`.

```{r prob01pval}
1 - pt(3.0621, 12)
```

# Problem 2

In 1899, biologist Hermon Bumpus presented as evidence of natural
selection a comparison of numerical characteristics of moribund house
sparrows that were collected after an uncommonly severe winter storm
and which had either perished or survived as a result of their
injuries. The data set `Bumpus.Rdata` contains the length of the
humerus (arm bone) in inches for 59 of these sparrows, grouped
according to whether they survived or perished. Analyze these data to
summarize the evidence that the distribution of humerus lengths
differs in the two populations.

## Solution

First, we can load Bumpus' data and see the `head`.

```{r prob02q}
load('./Bumpus.RData')
head(bumpus)
```

```{r mean02perished}
mean(bumpus[bumpus$Status == 'Perished',1])
sd(bumpus[bumpus$Status == 'Perished',1])
length(bumpus[bumpus$Status == 'Perished',1])
```

```{r mean02survived}
mean(bumpus[bumpus$Status == 'Survived',1])
sd(bumpus[bumpus$Status == 'Survived',1])
length(bumpus[bumpus$Status == 'Survived',1])
```

\[
{s}_{p} = \sqrt{\frac{23 \times 0.02354^2 + 34 \times 0.019839^2}{57}} = 0.02141
\]

\[
SE(\bar{Y}_2 - \bar{Y}_1) = 0.02141 \sqrt{\frac{1}{24} + \frac{1}{35}} = 0.005674
\]

\[
\text{d.f.} = 57
\]

Since, we cannot use the tables for this one, we have to rely on `R`.
Our null hypothesis is that ${H}_{0}: \mu_2 - \mu_1 = 0$ and,
therefore, $D = 0$.

\[
t = \frac{\bar{Y}_2 - \bar{Y}_1}{SE(\bar{Y}_2 - \bar{Y}_1)} = \frac{0.738 - 0.7279167}{0.005674}
= 1.777
\]

Therefore, the *p*-value can be calculated from `R`.

``` {r prob02p}
1 - pt(1.777, 57)
```

Since, *p*-value is $0.0404 < 0.05$, we can say with 95% confidence
that we reject our null hypothesis that the Humerus makes a difference
in the survival rate.

# Problem 3

Returning to Bumpus's data, obtain *p*-values from the *t*-test to
compare the humerus lengths for sparrows that survived and those that
perished, with and without the smallest length in the perished group
(length = 0.659 inches). Do the conclusions depend on this one
observation?

## Solution

First, we can load Bumpus' data and see the `head`.

```{r prob03q}
load('./Bumpus.RData')
bumpus <- bumpus[bumpus$Humerus != 0.659,]
head(bumpus)
```

```{r mean03perished}
mean(bumpus[bumpus$Status == 'Perished',1])
sd(bumpus[bumpus$Status == 'Perished',1])
length(bumpus[bumpus$Status == 'Perished',1])
```

```{r mean03survived}
mean(bumpus[bumpus$Status == 'Survived',1])
sd(bumpus[bumpus$Status == 'Survived',1])
length(bumpus[bumpus$Status == 'Survived',1])
```

\[
{s}_{p} = \sqrt{\frac{22 \times 0.0188195^2 + 34 \times 0.019839^2}{57}} = 0.0192735
\]

\[
SE(\bar{Y}_2 - \bar{Y}_1) = 0.0192735 \sqrt{\frac{1}{23} + \frac{1}{35}} = 0.0051734
\]

\[
\text{d.f.} = 56
\]

Since, we cannot use the tables for this one, we have to rely on `R`.
Our null hypothesis is that ${H}_{0}: \mu_2 - \mu_1 = 0$ and,
therefore, $D = 0$.

\[
t = \frac{\bar{Y}_2 - \bar{Y}_1}{SE(\bar{Y}_2 - \bar{Y}_1)} = \frac{0.738 - 0.730913}{0.0051734}
= 1.36989
\]

Therefore, the *p*-value can be calculated from `R`.

``` {r prob03p}
1 - pt(1.36989, 56)
```

In the previous part, our *p*-value was $0.0404 < 0.05$ which leads to
the fact that we would be rejecting our hypothesis that there is no
difference between the Hemurus of survived and perished with 95%
confidence. However, when we remove the influential point, we notice
that our confidence interval drops to 92%. Therefore, the conclusions
depend on this one observation.

# Problem 4

The following data are metabolic expenditures for eight patients
admitted to a hospital for reasons other than trauma and for seven
patients admitted for multiple fractures (trauma).

```{r prob04q, echo = FALSE}
ntp <- c(20.1, 22.9, 18.8, 20.9, 20.9, 22.7, 21.4, 20.0)
tp <- c(38.5, 25.8, 22.0, 23.0, 37.6, 30.0, 24.5, NA)
exp <- data.frame(ntp, tp)
namesp <- c("Non-trauma patients", "Trauma patients")
colnames(exp) <- namesp
library(knitr)
kable(exp, caption = "Metabolic Expenditures (kcal/kg/day)")
```

1. Determine the rank transformations for the data. *Hint*: A good way
   to proceed is to create a data set with one column for the
   metabolic expenditures and another column for the group (Trauma /
   Non-trauma). The function `rank` in R will be helpful to get the
   ranks.
2. Calculate the rank-sum statistic *T* by hand (taking the trauma
   patients to be group 1).
3. Calculate the *Z*-statistic $Z = \frac{T - \mu_T}{\sigma_T}$
4. Suppose we wanted to test the hypothesis ${H}_{0}: \eta_1 = \eta_2$
   vs. ${H}_{A}: \eta_1 > \eta_2$, where $\eta_1$ is the population
   median metabolic expenditure for trauma patients and $\eta_2$ is
   the population median metabolic expenditure for non-trauma
   patients. Find the one-sided *p*-value as the proportion of a
   standard normal distribution larger than the observed
   *Z*-statistic.
5. Use the `R` function `wilcox.text` to verify the rank-sum statistic
   and the *p*-value. *Hint*: Recall that R provides the statistic $W =
   T - \frac{\eta_1 (\eta_1 + 1)}{2}$. Also, you can use the option
   `correct = FALSE` to *not* use R's continuity correction, which should
   make the *p*-value calculated in this way identical to that calculated
   in part 4.
6. Use R to find a 95% confidence interval for the difference in
   population medians: the median metabolic expenditure for the
   population of trauma patients minus the median metabolic
   expenditure for the population of non-trauma patients. Interpret
   the interval in the context of the problem

## Solution

First, the data set has to be created.

```{r prob04s}
data <- c(20.1, 22.9, 18.8, 20.9, 20.9, 22.7, 21.4, 20.0,
          38.5, 25.8, 22.0, 23.0, 37.6, 30.0, 24.5)
group <- c("Non-trauma", "Non-trauma", "Non-trauma", "Non-trauma",
           "Non-trauma", "Non-trauma", "Non-trauma", "Non-trauma",
           "Trauma", "Trauma", "Trauma", "Trauma", "Trauma", "Trauma", "Trauma")
metabolic <- data.frame(data, group)
nonTrauma <- metabolic[metabolic$group == "Non-trauma",]
trauma <- metabolic[metabolic$group == "Trauma",]
head(metabolic)
```

Next, we calculate the medians of the two groups.

```{r prob04median}
summary(metabolic[metabolic$group == "Non-trauma",1])
summary(metabolic[metabolic$group == "Trauma",1])
```

### Rank transformation

First, we can sort the data in ascending order.

```{r prob04sort}
metabolic <- metabolic[order(metabolic$data),]
```

To do this, we use the `rank` function in `R`.

```{r prob04rank}
metabolic["rank"] <- rank(metabolic[,1])
metabolic
```

### Rank sum statistic

To perform the rank sum statistic, we have to add the sum of ranks
corresponding to each group. For the "Trauma" group,

\[
T = 7 + 10 + 11 + 12 + 13 + 14 + 15 = 82
\]

```{r prob04sum}
traumaSum <- colSums(Filter(is.numeric, metabolic[metabolic$group == "Trauma",]))
traumaSum
```

Then, we can see what is the number of combinations we need.

```{r prob04choose}
choose(length(metabolic[,1]), length(metabolic[metabolic$group == "Trauma",1]))
choose(length(metabolic[,1]), length(metabolic[metabolic$group == "Non-trauma",1]))
```

### *Z*-statistic

To calculate the *z*-statistic, we have to first calculate the mean and variance of the data.

\[
\mu_T = n_1 \bar{R} = 56
\]

```{r prob04mean}
meanT <- length(metabolic[metabolic$group == "Trauma",3])*mean(metabolic[,3])
meanT
```

Similarly, we can calculate the standard deviation of the T statistic.

\[
\sigma_T = {s}_{R} \sqrt{\frac{{n}_{1} {n}_{2}}{{n}_{1} + {n}_{2}}} = 8.633269
\]

```{r prob04sd}
sigT <- sd(metabolic[,3]) *
    sqrt((length(metabolic[metabolic$group == "Trauma",3]) *
          length(metabolic[metabolic$group == "Non-trauma",3])) /
         (length(metabolic[metabolic$group == "Trauma",3]) +
          length(metabolic[metabolic$group == "Non-trauma",3])))
sigT
```

Finally,

```{r prob04z}
z <- (traumaSum[2] - meanT)/sigT
z
```

\[
z = \frac{T - {\mu}_{T}}{\sigma_T} = \frac{82 - 120}{8.633269}
\]

\[
\implies z = 3.01
\]

### Hypothesis testing

The hypotheses for this problem is given below

\[
{H}_{0}: \eta_1 = \eta_2
\]

\[
{H}_{A}: \eta_1 > \eta_2
\]

The p value is derived from the *Z*-statistic.

\[
p = P[z \geq Z] = 1 - P[z < Z] = 1 - P[z < 3.01]
\]

```{r prob04pval}
1 - pnorm(3.01)
```

\[
\implies p = 0.0013
\]

### Wilcox test

We know that

\[
W = T - \frac{\eta_1(\eta_1 + 1)}{2} = 82 - \frac{7 (7 + 1)}{2} = 82 - 28 = 54
\]

`R` has the function `wilcox.test` for this purpose.

```{r prob04wilcox}
wilcox.test(metabolic[metabolic$group == "Trauma",1],
            metabolic[metabolic$group == "Non-trauma",1],
            alternative="greater", correct=FALSE,
            conf.int=TRUE, conf.level=0.95)
```

### Confidence of interval

```{r prob04ci}
wilcox.test(metabolic[metabolic$group == "Trauma",1],
            metabolic[metabolic$group == "Non-trauma",1],
            alternative="two.sided", correct=FALSE,
            conf.int=TRUE, conf.level=0.95)
```

Therefore, the 95% confidence interval is (2, 16.2). The variance of
the median seems to be high. This is due to the fact that the sample
size is too small. Having a larger sample size will give much more
accurate results.
